Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-05 model.tex
editing

Delay-coordinate embedding allows one to reconstruct a dynamical system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(X) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
\]
from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of this modeling approach.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92} to estimate $m$.  A
plot of the data from Figure~\ref{fig:ipc}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
\begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/svd53dipc.png}
    \caption{\svdfive}
    \label{fig:svdfiveEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure}%

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{[[{\color{red}Joshua:Perhaps add a similar graphic of \gcc to show contrast}{\color{green}agreed I will work on this}]]A 3D projection of a delay-coordinate embedding of the trace
 from Figure~\ref{fig:ipc} with a delay ($\tau$) of 100,000 instructions.
 }
 \label{fig:embedding}
 \end{figure}

[[Update this paragraph if we include the second plot with \gcc]]
%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \col instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism%\footnote{A deeper analysis of
  %Figure~\ref{fig:embedding}---as alluded to on the previous
  %page---supports that diagnosis, confirming the presence of a chaotic
  %attractor in these cache-miss dynamics, with largest Lyapunov
  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  %12-dimensional reconstruction space \cite{mytkowicz09}.}
  .  That structure can be used to build a forecast model. 
 \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
shown in Figure~\ref{fig:lmacol}.
%
\begin{figure}[htbp]
  \centering
   \includegraphics[width=\textwidth]{figs/colPredShortTS}
    \caption{A forecast of the last 4,000 points of the signal in
      Figure~\ref{fig:ipc} using an LMA-based strategy on the
      embedding in Figure~\ref{fig:embedding}.  Red circles and blue
      $\times$s are the true and predicted values, respectively;
      vertical bars show where these values differ.}
\label{fig:lmacol}
\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor load (ipc) of the \gcc program from the SPEC cpu2006 benchmark suite, the
prediction is far less accurate; see Figure~\ref{fig:lmagcc}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{figs/colPredShortTS}
     \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of the last 4,000 points of a
       processor-load performance trace from the \gcc
       benchmark.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}
\label{fig:lmagcc}
\end{figure}

This gets at the utility of the contribution of this paper: If these traces both come from deterministic
systems---computers---then why are they not equally predictable?  Our
conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum. If this is the case then \gcc might be much better predicted using a method like Random Walk or \naive which can effectively predict in the presence of complexity. This is explored more rigorously in the results Section of this paper (Section \ref{sec:results}. So that we don't have to compare the forecast accuarcy by visual comparing the predictions we calculate a figure of merit to compare the predictions.

@@

\begin{figure}[htbp]
  \centering
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colMeanForecast}
    \caption{\col na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccMeanForecast}
    \caption{\gcc na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%      
  \\
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
    \caption{\col ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
    \caption{\gcc ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \\
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colLMAForecast}
    \caption{\col LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
    \caption{\gcc LMA}
    \label{fig:gccLMA}
  \end{subfigure}

  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{
For each of these, we plot the predicted value $p_n$ against the correct value $c_n$. On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction. }\label{fig:gcc_vs_col}  
  %\end{subfigure}
\end{figure} 
In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.

>>>>

Delay-coordinate embedding allows one to reconstruct a dynamical system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(\vec{X}) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
\]
from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of this modeling approach.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92} to estimate $m$.  A
plot of the data from Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following this
procedure, are shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
\begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  
    \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure} 
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/svd53dipc2.png}
    \caption{\svdfive}
    \label{fig:svdfiveEmbedding}
  \end{subfigure}%   
%

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{A 3D projection of the delay-coordinate embeddings of the traces
 from: (a) \col seen in Figure~\ref{fig:col-ts} (b)\gcc seen in Figure~\ref{fig:gcc-ts} and (c) \svdfive seen as the fifth regime in Figure~\ref{fig:svd-ts-colored}.
 }
 \label{fig:embedding}
 \end{figure}

%% can cut for space if need be:
The coordinates of each point in these plots are differently delayed
elements of the instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding} (a) is an indication of
structure in the information transmission process, that can be leveraged for prediction. Where the water becomes muddy is with embedding like those in Figure~\ref{fig:embedding} (b) \& (c). Figure~\ref{fig:embedding} (b) shows little to no structure, this embedding is essentially a noise ball, with an embedding like this we would expect the forecast accuracy that uses this structure to be lacking. However, with Figure~\ref{fig:embedding} (c) more structure is present but the structure is surrounded in noise. Is the structure enough to aid in accurately forecasting the signal? This is a difficult question to answer and one that comes back to the main contributions of this paper. It may be that \gcc has structure but structure that is not amenable by this method, similarly \svdfive clearly has structure but it is obfuscated by noise. By analyzing qualitatively the amount of complexity in these signals we can answer questions like this. If enough predictive structure is present in these embeddings as seen in Figure~\ref{fig:embedding} (a) and possibly in (c), that structure can be used to build a forecast model which we discuss in the next section.
 \subsubsection{LMA: Using dynamics in forecasting {\color{blue} EDITABLE}}

Given a nonlinear model of a deterministic dynamical system in the
form of a delay-coordinate embedding like Figure~\ref{fig:embedding},
one can build deterministic forecast algorithms by capturing and
exploiting the geometry of the embedding.  Many techniques have been
developed by the dynamical systems community for this purpose
(e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).  Perhaps the most straightforward
is the ``Lorenz method of analogues'' (LMA), which is essentially
nearest-neighbor prediction in the embedded state
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
forecast---works quite well on the trace in Figure~\ref{fig:ipc}, as
shown in Figure~\ref{fig:lmacol}. Visually distinguishing if Figure~\ref{fig:lmasvdfive} is a good prediction is hard. To more fairly compare these forecasts as well as the forecasts made by the other three prediction strategies we calculate a figure of merit, which we introduce in Section \ref{sec:accuracy}.
%
\begin{figure}
   \centering
\begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colPredShortTS}
    \caption{\col }
    \label{fig:lmacol}
  \end{subfigure}%  
    \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colPredShortTS}
    \caption{\gcc}
    \label{fig:lmagcc}
  \end{subfigure} 
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colPredShortTS}
    \caption{\svdfive}
    \label{fig:lmasvdfive}
  \end{subfigure}% 
  \caption{{\color{red} [[actually need to generate this figure]]}An LMA-based forecast of
       processor-efficiency performance trace from the\col ,\gcc and \svdfive generating processes.  Red circles and blue $\times$s are the true and
       predicted values, respectively; vertical bars show where these
       values differ.}\label{fig:lmapredictions}
 \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{figs/colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{figs/colPredShortTS}
%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
On the other hand, if we use the same approach to forecast the
processor efficiency (IPC) of the \gcc time series, the
prediction is far less accurate; see Figure~\ref{fig:lmagcc}.

This gets at the utility of the contribution of this paper: These time series all come from the same system---computers---but they are not equally predictable (at this point by LMA). For a practitioner is it the case that

  Our
conjecture is that while they come from similar systems \gcc produces processor load traces on the top of the complexity spectrum whereas \col produces processor loads with complexity in the mid to low region of the complexity spectrum. If this is the case then \gcc might be much better predicted using a method like Random Walk or \naive which can effectively predict in the presence of complexity. This is explored more rigorously in the results Section of this paper (Section \ref{sec:results}. So that we don't have to compare the forecast accuarcy by visual comparing the predictions we calculate a figure of merit to compare the predictions.

@@

\begin{figure}[htbp]
  \centering
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colMeanForecast}
    \caption{\col na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccMeanForecast}
    \caption{\gcc na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%      
  \\
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colARIMAForecast}
    \caption{\col ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccARIMAForecast}
    \caption{\gcc ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \\
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colLMAForecast}
    \caption{\col LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/gccLMAForecast}
    \caption{\gcc LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{
{\color{red}[[Joshua: I should probably add \svdfive as it is a persistent example making this figure 9x9.}For each of these, we plot the predicted value $p_n$ against the correct value $c_n$. On this type of plot a perfect prediction lies exactly on the diagonal, that is the line $p_n = c_n$, e.g., \ref{fig:colLMA} is a near perfect prediction where-as \ref{fig:gccARIMA} is a very poor prediction. }\label{fig:gcc_vs_col}  
  %\end{subfigure}
\end{figure} 
In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.
